A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network

### 2018-10-31 anzjs2018/anzjs2018.tex
some words + graphs
More edits

% \doublespace

@@

\begin{document}
\cleanlookdateon
\runningheads{Modeling transit vehicles in real-time}{TOM~ELLIOTT AND THOMAS~LUMLEY}
\title{A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network}
\author{Tom Elliott\corrauth~and~Thomas~Lumley}
\affiliation{University of Auckland}
\address{
    Department of Statistics, University of Auckland, 
    Auckland 1142, New Zealand\\
    Email: \texttt{tom.elliott@auckland.ac.nz}
}

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherance,
but also information about traffic conditions along the remainder of its route.
Road networks are dynamic, 
and can change from free-flowing to highly congested very quickly,
which can effect the arrival time of transit vehicles, 
buses in particular,
% This in turn leads to commuter frustration when they have to wait longer than
% initially anticipated,
so any reliable prediction framework needs to account for \rt traffic conditions.
In this paper,
we propose a particle filter that can be used to estimate travel times
along roads within a transit network based solely on GTFS data.
Due to the \rt nature of the problem and computational demands of such an approach,
we demonstrate the faesibility of our program, 
written in the R programming language and making extensive use of C++.
The application provides the basis from which future work will be based,
most notably for combining the \rt travel times with historical data and a 
formal road network model to make reliable arrival time predictions.
% Of course, the major constraint on any prediction framework is going to be
% computational efficiency.
% Due to the realtime nature of the problem, predictions need to be 
% available as soon as possible after observing the vehicles' positions,
% which are updated approximately every 30~seconds.
% In this paper, we describe our C++ implementation of a particle filter to estimate
% vehicle state in combination with a network state model,
% allowing us to model transit vehicle flow through the network.
Development and testing has taken place here in Auckland, New Zealand,
although we expect our methods are general enough to be applied to any
transit network using GTFS.
\end{abstract}

>>>>

\doublespace

@@

\begin{document}
\cleanlookdateon
\runningheads{Modeling transit vehicles in real-time}{TOM~ELLIOTT AND THOMAS~LUMLEY}
\title{A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network}
\author{Tom Elliott\corrauth~and~Thomas~Lumley}
\affiliation{University of Auckland}
\address{
    Department of Statistics, University of Auckland, 
    Auckland 1142, New Zealand\\
    Email: \texttt{tom.elliott@auckland.ac.nz}
}
\ack{
    We would like to thank Auckland Transport for providing free,
    open access to the \rt data; and the Center for e-Research at the
    University of Auckland for providing and maintaining the virtual
    machine used in our work.
}

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherance,
but also of traffic conditions along the remainder of the route.
Road networks are dynamic, 
and can change from free-flowing to highly congested very quickly,
which impacts the arrival time of transit vehicles,
particularly buses which often share the road with other vehicles,
so reliable predictions need to account for \rt and future traffic conditions.
The first step in this process is to construct a framework with which road state
(traffic conditions) can be estimated using \rt transit vehicle position data.
We propose a framework which implements a vehicle model using a particle filter
to estimate travel times along roads,
used by a second model to estimate \rt traffic conditions which can
later be used for arrival time prediction.
We demonstrate the \rt faesibility of such an approach in \rt,
and detail our implementation in R and C++.
Development and testing took place in Auckland, New Zealand,
with future work planned to incorporate historical data into the network model
to improve the reliability of ETAs.
\end{abstract}

### 2018-10-31 anzjs2018/sections/01_introduction.tex
More edits

In public transport, \rt information (RTI), for example estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and in some places make decisions on the best route to take.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
One of these is inaccurate ETAs,
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or  because the schedule is poorly calibrated.
Another is that buses are shown as \emph{on-time} if they are not present
in the \rt system,
no matter whether this is due to technical problems,
or that the bus has failed to show up;
in this case, schedule times are used to determine ETAs and, 
if the bus doesn't show up on time,
the service is removed from the \rt board 
and we get what passengers have described as ``Ghost Buses''.

Arrival time prediction is only as accurate as the underlying model,
and while a lot of work has gone into developing public transport vehicle models
\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
in many public transport systems no formal model is used.
Instead, ETAs are solely based on the scheduled arrival time
adjusted by the vehicle's delay at the most recent stop, if available.
This assumes that the schedule is valid,
and that there is no unusual congestion along the route,
neither of which are valid assumptions,
particularly in our test area of Auckland, New Zealand,
where infrastructure for buses (such as priority lanes) is limited.
This behaviour is exaggerated when there is a large distance between stops,
as the ETA is only updated at stops so there can be a long time between updates.

To avoid these problems and improve the reliability of ETAs,
we required two things:
first, a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data;
and secondly, a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modeling approaches were explored, 
such as the Kalman filter (used by \cite{Dailey_2001} and \cite{Cathey_2003}, ...),
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter approach demonstrated
by \cite{Hans_2015}.
The last of these, the particle filter, was chosen for several reasons,
discussed in Section~\ref{sec:pf}.

In order to improve the reliability of ETAs,
congestion along intermediate roads should be incorporated into the predictive model.
One solution is to use the travel time of preceeding buses along the same roads
to predict travel times for the current bus,
such as proposed by \cite{Yu_2010},
who were able to improve prediction accuracy
by using the travel time of vehicles on the same \emph{route}.
However, this only works well for high-frequency routes, so
\cite{Yu_2011} modeled travel times along roads where several major routes converged,
which showed further improvement however is limited in scope.
In general, however, no comprehensive network modeling approach has been proposed us
solely GPS position data to estimate arrival times
while accounting for congestion.

In this paper, we describe a \rt vehicle and network state
model which uses the vehicles traveling throughout the transit road network
to update congestion information in \rt.
As the vehicle model is implemented using a particle filter,
we demonstrate its faesibility as a valid real-time option.
From the particle filter estimates of vehicle trajectories,  
The particle filter provides estimates of vehicle travel times along roads,
which can be used to update the road network's state.
Future work involves using this travel time information 
in combination with historical data to develop an arrival time prediction model.

>>>>

In public transport, \rt information (RTI), such as estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and potentially make decisions on the best route to take.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
One of these is inaccurate ETAs,
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or  because the schedule is poorly calibrated.
Another is that buses are shown as \emph{on-time} if they are not present
in the \rt system,
regardless of whether this is due to technical problems,
or the bus failing to show up;
in either case, schedule times are used to determine ETAs and, 
if the bus doesn't show up on time,
the service is removed from the \rt board 
and we get what passengers have described as ``Ghost Buses''.

Arrival time prediction is only as reliable as the underlying model,
and while a lot of work has gone into developing public transport vehicle models
\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
in many public transport systems no formal model is used.
Instead, ETAs are solely based on the scheduled arrival time
adjusted by the vehicle's delay at the most recent stop, if available.
This assumes that the schedule is valid,
and that there is no unusual congestion along the route,
neither of which are valid assumptions,
particularly in our test area of Auckland, New Zealand,
where infrastructure for buses (such as priority lanes) is limited.
This behaviour is exaggerated when there is a large distance between stops,
as the ETA is only updated at stops so there can be a long time between updates.

To avoid the problems associated with the arrival time predictions
based solely on reported delays,
a more robus modeling and prediction framework is required.
This should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data;
and a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modeling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter \citep{Hans_2015}.
The last of these was chosen for several reasons
which are discussed in Section~\ref{sec:pf}.

The next step is to construct a method of estimating traffic conditions.
This can be done by using the travel time of preceeding buses along the road
to predict travel times for the current bus.
\cite{Yu_2010} obtained improved prediction accuracy by using travel times
from preceeding buses servicing the same route.
\cite{Hans_2015} used a similar approach,
instead referring to headway, the time between consecutive vehicles
at a point on the route.
However, this approach only works well for high frequency routes.
\cite{Yu_2011} showed further improvements by combining travel times 
from several routes which converged,
but is limited to predefined, converging routes.
In general, however, no comprehensive network modeling approach has been proposed using
solely GPS position data to accounting for congestion when estimating arrival times.
We describe a simple method of constructing a road network
from GTFS data (section~\ref{sec:gtfs}),
which allows all transit vehicles traveling along a road to contribute to tis state,
regardless of which route they are serving.

In this paper, we describe a \rt vehicle and network state
model which uses the vehicles traveling throughout the transit road network
to update congestion information in real-time.
As the vehicle model is implemented using a particle filter,
a computationally intensive method,
we demonstrate its faesibility as a valid real-time option.
The vehicle model provides estimates of travel times along roads,
which can be used to update the road network's state.
Future work is planned to use this travel time information 
in combination with historical data to develop an arrival time prediction model.

### 2018-10-31 anzjs2018/sections/02_gtfs.tex
More edits

As mentioned in Section~\ref{sec:intro},
there are some major issues with the current prediction method in Auckland.
These are almost directly attributed to GTFS-realtime trip updates,
which are currently the sole source of data for ETAs.
Trip updates are reported whenever a vehicle arrives or departs a stop,
and includes the delay between the scheduled and actual arrival times,
and is propagated to all future stops to adjust their ETAs.
This assumes that the schedule is well calibrated and the time between scheduled arrivals
is representative of the real-world travel time between stops. 
In many places in Auckland, 
traffic congestion is highly dynamic and changes from day to day,
so even a well-calibrated schedule is an unreliable baseline.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable,
at least directly,
so here we construct a transit road network consisting of intersections
and the roads between them.
By generalising to physical roads,
we remove any route-dependency,
which increases the quantity of data available for each road.
Then, each route can instead be represented as a sequence of road segments,
each with a travel time state.

To construct the network,
we use stops as nodes, so that
routes with a common sub-sequence of stops will share the roads between them;
roads are then defined as the links between stops,
as demonstrated in Figure~\ref{fig:network_creation}.
In section~\ref{sec:kf} we present a model for estimating vehicle travel time
along a road segment.
There are several places where this doesn't quite work, 
for example express routes which do not service certain stops,
and multi-stop locations,
which results in overlapping road segments.
Future work will look at using intersection points of the route shapes
as nodes in the network to improve the coverage and reduce overlap.

@@

One important consideration regarding the Auckland Transport's \rt implementation is that
buses are programmed to report their location when arriving at or departing from
bus stops, as well as some major intersections.
To further complicate this,
these positions can be preemtive,
(i.e., the bus is almost there, but not quite),
so consecutive observations can show a bus traveling backwards.
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state become continuing.
This way, we avoid degeneration of the particle fitler (Section~\ref{sec:model_perf})
at the cost of losing one observation that is most likely invalid anyway.

>>>>

As mentioned in Section~\ref{sec:intro},
there are some major issues with the current prediction method in Auckland.
These are almost directly attributed to GTFS-realtime trip updates,
which are currently the sole source of data for ETAs.
Trip updates are reported whenever a vehicle arrives or departs a stop,
and includes the delay between the scheduled and actual arrival times,
and is propagated to all future stops to adjust their ETAs.
This assumes that the schedule is well calibrated and the time between scheduled arrivals
is representative of the real-world travel time between stops. 
Therefore, an alternative estimation approach using \rt travel times 
between stops is more adequate.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable,
at least directly,
so here we construct a transit road network consisting of intersections
and the roads between them.
By generalising to physical roads,
we remove any route-dependency,
which increases the quantity of data available for each road.
Each route is then represeted as a sequence of road segments,
each with an associated travel time.

The transit network is constructed by splitting each route
into spatially identifiable segments,
each representative of a physical road.
The simplest way to do this is to use bus stops as nodes in the network
(Figure~\ref{fig:network_creation_1}),
and connecting roads as edges.
In this way, routes that serice the same subsequence of stops
will all contribute to congestion information for the connecting roads,
as shown in Figure~\ref{fig:network_creation_2}.
There are several drawbacks to this method,
for example routes merge at physical intersections, not bus stops,
so some overlap will occur as shown.
However, this is minimal, and our simplification is sufficent for the current work.

@@

One important consideration regarding Auckland Transport's \rt implementation is that
buses are programmed to report their location when arriving at or departing from
bus stops, as well as some major intersections.
To further complicate this,
these positions can be pre-emtive,
(i.e., the bus is almost there, but not quite),
so consecutive observations can show a bus traveling backwards.
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state before continuing.
This way, we avoid degeneration of the particle fitler (Section~\ref{sec:model_perf})
at the cost of losing one observation that is most likely invalid anyway.

### 2018-10-31 anzjs2018/sections/03_models.tex
More edits

Real-time tracking applications such as ours often use
a Bayesian filtering approach,
or recursive Bayesian estimation,
in which the data is processed sequentially instead of all at once,
reducing computational demands.
These models assume an underlying Markov process with state $\mathcal{X}_k$
and system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
of which observations $\mathcal{Y}_k$ are made
with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
giving the following general model:
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
The transition function $f$ describes the relationship between consecutive states,
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

@@

We chose to employ a particle filter to estimate $\bX_k$,
due to do its high flexibility and use in recent 
transit vehicle modeling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

@@

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size 
\begin{equation}
\label{eq:neff}
N_{\text{eff}} = \frac{1}{\sum_i (W_k^{(i)})^2}
\end{equation}
falls below a specified threshold $N_{\text{thres}}$.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

@@

The particle filter allows us to develop a model of several important bus behaviours:
\begin{itemize}
\item non-constant speed along roads,
\item optional stopping and waiting at bus stops while passengers board and disembark, and
\item optional stopping and waiting at intersections or traffic lights (future work).
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's acceleration changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on the values used by \cite{Hans_2015};
future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.

The stop positions are represented by their distance into trip,
$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$.
The $stop\_index$ function returns the index of a stop,
given a distance traveled and set of stop distances.
The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
which implements those features discussed above.
Lines~2--9 are necessary to decrease the degeneration rate of the filter,
which can occur when there is a long delay between observations
simply because the bus has not moved due to traffic or other reasons. 
We therefore place positive probability for some particles to remain stationary.

Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
which involves changing the vehicle's speed each second (i.e., acceleration).
Lines~13--25 handle arrival behaviour at stops:
if the particle passes the next stop,
it has a chance of stopping (lines~18--19),
in which case the dwell time is sampled and deducted from the 
time remaining to travel (lines~20--21).
Finally, to track travel time along segments,
at each point where the vehicle travels (lines 9 and 26),
the travel time for that segment, $z_j$, is incremented.
The mutation ends once the travel time $\delta$ reaches zero,
or when the vehicle reaches the end of the route (lines 10, 14).
Once finished, the particle's new state is returned (line 29).

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection, \cite{Snyder_1998}),
allow the placement of a bivate normal likelihood on the data.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
Now, the geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
cane be calculated using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible due to the discretization of the state provided by the particle filter;
in other approaches, such as the Kalman filter,
a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance traveled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

@@

The primary objective of the network model is to estimate the \rt traffic conditions
(travel time) along roads in the transit network, 
as well as make short-term predictions for estimating arrival times.
In this paper, we develop simple model in which each road is treated independently.
Future work will investigate using correlations with adjacent roads and historical data
to improve the model and subsequent predictions.

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$,
and is assumed to be constant with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
Observations of the state, $Z_c^\ell$, are obtained from the particle filter in Section~\ref{sec:pf},
and have measurement error $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ as estimated 
using uncertainty of particle estimates.
The model from (\ref{eq:rbe_model}) now reduces to
\begin{equation}
\begin{split}
\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation}
where $\Delta_c = t_c - t_{c-1}$.

Since multiple vehilces can travel along a road at the same time,
we have used an information filter to implement the network model.
The information filter is a transformation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
This allows multiple observations to be added together to update the state
in a single iteration.

The state is estimated by its mean $\hat \theta_c^\ell = \mathrm{E}(\theta_c^\ell)$
and uncertainty $P_c^\ell = \mathrm{var}(\theta_c^\ell)$,
which are predicted from the previous state estimate using the prediction
equations
\begin{align*}
\label{eq:kf_transition}
\hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c|c-1} \\
P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + \Delta_c^\ell \nu_c^\ell
\end{align*}

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c = (P_{c|c-1})^{-1}$
and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
The travel time estimate of vehicle $m$ along segment $\ell$
is $\bar z_c^{m\ell}$, with uncertainty $s^{m\ell}_c$,
and is transformed to a measurement information covariance matrix 
$B_c^{m\ell} = (s^{m\ell}_c)^{-2}$
and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
The total information is the sum of the information over all $M$ vehicles
that traversed segment $\ell$ since the last update,
so the update equation is
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}.
\end{align*}
The desired parameter estimates (tavel time) are obtained 
by the inverse transformations
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}
\end{equation*}
which can now be used to predict the travel times of upcoming vehicles.

>>>>

Real-time tracking applications often use
a Bayesian filtering approach,
or recursive Bayesian estimation,
in which the data is processed sequentially instead of all at once,
reducing computational demands
\citep{Anderson_2012}.
These models assume an underlying Markov process with state $\mathcal{X}_k$
and system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
of which observations $\mathcal{Y}_k$ are made
with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
giving the following general model:
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
The transition function $f$ describes the relationship between consecutive states,
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

@@

We use a particle filter to estimate $\bX_k$,
due to do its high flexibility and use in recent 
transit vehicle modeling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function, 
discussed in section~\ref{sec:pf_update}.
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

@@

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size 
\begin{equation}
\label{eq:neff}
N_{\text{eff}} = \frac{1}{\sum_i (W_k^{(i)})^2}
\end{equation}
falls below a specified threshold $N_{\text{thres}}$
\citep{}.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

@@

The particle filter allows us to model several important bus behaviours:
\begin{itemize}
\item non-constant speed along roads,
\item optional stopping and waiting at bus stops while passengers board and disembark, and
\item optional stopping and waiting at intersections or traffic lights (future work).
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's speed changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on the values used by \cite{Hans_2015};
future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.

The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
which implements those features discussed above.
The stop positions are represented by their distance into trip,
$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$.
The $stop\_index$ function returns the index of a stop,
given a distance traveled and set of stop distances.
Lines~2--9 are necessary to decrease the degeneration rate of the filter,
which can occur when there is a long delay between observations
simply because the bus has not moved due to traffic or other reasons. 
We therefore place positive probability for some particles to remain stationary.

Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
which involves changing the vehicle's speed each second (i.e., acceleration).
Lines~13--25 handle arrival behaviour at stops:
if the particle passes the next stop,
it has a chance of stopping (lines~18--19),
in which case the dwell time is sampled and subtracted from the 
time remaining to travel (lines~20--21).
Finally, to track travel time along segments,
at each point where the vehicle travels (lines 9 and 26),
the travel time for the current segment, $z_j$, is incremented.
The mutation ends once the travel time $\delta$ reaches zero,
or when the vehicle reaches the end of the route (lines 10, 14).
Once complete, the particle's new state is returned (line 29).

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection, \cite{Snyder_1998}),
allow the placement of a bivate normal likelihood on the data.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
The geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
is calculated using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible due to the discretization of the state provided by the particle filter;
in other approaches, such as the Kalman filter,
a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance traveled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

@@

The primary objective of the network model is to estimate the \rt traffic conditions
(travel time) along roads in the transit network, 
as well as make short-term predictions for estimating arrival times.
In this paper, we develop a simple model in which each road is treated independently.
Future work will explore correlations between adjacent roads and using historical data
to improve the model and subsequent predictions.

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$,
and is assumed to be constant over time 
with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
Observations of the state, $Z_c^\ell$, are obtained from the particle filter in Section~\ref{sec:pf},
and have measurement error $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ as estimated 
using the uncertainty of particle estimates.
The model from (\ref{eq:rbe_model}) now reduces to
\begin{equation}
\begin{split}
\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation}
where $\Delta_c = t_c - t_{c-1}$.

Since multiple vehicles can travel along a road simultaneously,
we have used an information filter to implement the network model.
The information filter is a transformation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
This allows multiple observations to be added together to update the state
in a single iteration.

The state is estimated by its mean $\hat \theta_c^\ell = \mathrm{E}(\theta_c^\ell)$
and uncertainty $P_c^\ell = \mathrm{var}(\theta_c^\ell)$,
which are predicted from the previous state estimate using the predictive model
\begin{align*}
\label{eq:kf_transition}
\hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c|c-1} \\
P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + \Delta_c^\ell \nu_c^\ell
\end{align*}

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c = (P_{c|c-1})^{-1}$
and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
The travel time estimate of vehicle $m$ along segment $\ell$
is $\bar z_c^{m\ell}$, with uncertainty $s^{m\ell}_c$,
and is transformed to a measurement information covariance matrix 
$B_c^{m\ell}~=~(s^{m\ell}_c)^{-2}$
and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
The total information is the sum of the information over all $M_c^\ell$ vehicles
that traversed segment $\ell$ since the last update,
so the update equation is
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} b^{m\ell}_{c}.
\end{align*}
The desired parameter estimates are obtained 
by the inverse transformations
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}
\end{equation*}
which can now be used to predict the travel times of upcoming vehicles.
